Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization

### 2013-10-03 abstract.tex
Theorems for GC and updated abstract.

Pushdown analysis is better than finite-state analysis in precision and
performance. Why then have we not seen total widespread adoption of these
techniques? For one, the known techniques are technically burdened and
difficult to understand or extend. Control structure of the programming language
gets pulled into the model of computation, which makes extensions to
non-pushdown control structures, such as \rackett{call/cc} or \rackett{shift} and
\rackett{reset}, non-trivial.

We show a derivational approach to abstract interpretation that yields
transparently sound static analyses that can precisely match calls and
returns when applied to well-known abstract machines. We show that
adding memoization and segmenting the continuation into bounded pieces
leads to machines that abstract to static analyses for context-free
reachability by simply bounding the stores. This technique allows us
to derive existing, more technically involved analyses, and a novel
pushdown analysis for delimited, composable control.

>>>>

Pushdown analysis improves upon finite-state analysis in precision and performance since it can rule out spurious paths caused by improperly matched calls and returns.
%
Despite this, we have not seen total widespread adoption of these techniques for analyzing higher-order languages.
%
A possible explanation is that the known techniques are technically burdened and difficult to understand or extend.
%
Control structure of the programming language gets pulled into the model of computation, which makes extensions to non-pushdown control structures, such as first-class control operators, non-trivial.
%%

%%
We show a derivational approach to abstract interpretation that yields transparently sound static analyses that can precisely match calls and returns when applied to well-known abstract machines.
%
We show that the abstraction techniques in the literature have an intuitive operational meaning, even in the concrete setting, that build off well-known programming techniques such as memoization.
%
This approach allows us to derive the existing, more technically involved analyses, and a novel pushdown analysis for delimited, composable control.

### 2013-10-03 content.tex
Theorems for GC and updated abstract.

The $\Addr$ space is what controls the precision of the model.
%
For a concrete semantics, we require the allocation meta-function $\alloc :
\State \to \Addr$ to return fresh addresses ($\maddr \notin \dom(\mstore)$), but any choice of address is sound.
%
If $\alloc$ only uses addresses from a finite subset of $\Addr$, then the state space without continuations is finite, and the space of contination frames is finite (thus the pushdown system interpretation is apt).
%
This meta-function along with the commitment to having no recursive data-structures is the key to the AAM technique.
%
All recursion can be expressed by indirecting through addresses into a store --- tying Landin's knot.
%
We borrow this technique on top of our own.

@@

To turn this semantics into PDCFA's algorithm for constructing a Dyck state graph, given the appropriate $\alloc$, we apply a widening operator to make $\mstore$, $\mkont$, and $\mmemo$ shared amongst all states (meaning states then become represented without these compontents), in what is then called a $\System$.
%
Since these components are shared, they are always the least upper bound of any respective component that the semantics produces (the intermediate set $I$) in order to stay sound.
%
We use a meta-function $\wn$, ``wide to narrow'', to shift between the two representations of states.

%% FIXME: remove stores from memo-table since they will always be unchanged
%% TODO: Show that store widening and store/memo/KTab widening have a bisimulation.
\begin{align*}
  \mastate \in \sa{State} &= (\Expr \times \Env) \times \Kont \\
  \System &= \wp(\sa{State} \times \Store) \times \wp(\sa{State}) \times \Store \times \KTab \times \Memo \\
  {\mathcal F} &: \System \to \System \\
  {\mathcal F}(S, F, \mstore, \mktab, \mmemo) &= (S \cup S', F', \mstore', \mktab', \mmemo') \\
  \text{where } I &= \setbuild{\mstate'}{\mastate \in F, \wn(\mastate,\mstore,\mktab,\mmemo) \stepto \mstate'} \\
                \mstore' &= \bigsqcup\setbuild{\mstore'}{\wn(\_, \mstore',\_,\_) \in I} \\
                \mktab' &=  \bigsqcup\setbuild{\mktab'}{\wn(\_, \_, \mktab', \_) \in I} \\
                \mmemo' &=  \bigsqcup\setbuild{\mmemo'}{\wn(\_, \_, \_, \mmemo') \in I} \\
                S' &= \setbuild{(\mastate',\mstore')}{\wn(\mastate', \_, \_, \_) \in I} \\
                F' &= \setbuild{\mastate'}{(\mastate',\_) \in S' \setminus S} \\
                \wn(\tpl{(\mexpr,\menv),\mkont},\mstore,\mktab,\mmemo)
                   &= \tpl{(\mexpr,\menv),\mstore,\mkont,\mktab,\mmemo}
\end{align*}

@@

The different tables we have encode information about execution history that we prove is invariant.
\begin{align*}
  \inv(\tpl{\mpoint, \mstore, \mkont, \mktab, \mmemo}) &=
          \dom(\mmemo) \subseteq \dom(\mktab) \\
  &\wedge \forall \mctx\equiv(\mpoint',\mstore') \in \dom(\mktab),
                  \mkont^\circ \in \mktab(\mctx),
                  \mkont' \in \unroll{\mktab}{\mkont^\circ}. \\
  &\qquad \exists \mkont^* \in \unroll{\mktab}{\mkont}.
            \tpl{\mpoint', \mstore', \mkont'} \stepto^* \tpl{\mpoint, \mstore, \mkont^*} \\
  &\qquad \wedge \forall (\mval, \mstore') \in \mmemo(\mctx).
            \tpl{\mpoint', \mstore', \mkont'} \stepto^* \tpl{\mval,\mstore', \mkont'}
\end{align*}

The first is that the memo table only contains information about previously seen contexts; we need this to infer that there was at least one call leading to the memoized context such that we can use stack irrelevance to justify skipping to the memoized result.
%
The second (the existential clause) states that all the calling contexts in the continuation table reach some unrolling of the current state.
%
The final invariant states that the memo table is correct: all calling contexts reach the memoized result, with well-bracketing.

@@

This semantics of course enjoys the soundness properties that GC does not introduce dangling pointers, and (in the concrete) any two reduction sequences of the same length lead to states equal up to GC.
%
Unconstrained allocation functions

@@

\begin{align*}
  \mctx \in \Context &::= ((\mexpr,\menv),\mstore) \mid (\vcomp{\mkont},\mval,\mstore) \mid \msctx \\
  \msctx \in \SContext &::= ((\mexpr,\menv),\maddr) \\
  \mskont \in \SKont &= \krt{\msctx} \mid \ldots \\
  \mval \in \Value &= \vclo{\slam{\mvar}{\mexpr}}{\menv} \mid \vcomp{\mskont} \\
  \mmkont \in \MKont &::= \kmt \mid \kprompt{\mctx} \\
  \mktab \in \KTab &= (\Context \to \mathit{KTabRange}) \\
                   &\cup (\Addr \to \wp(\Store))
\end{align*}

@@

At return boundaries, the memo table must add the result to all the represented contexts:=
\begin{align*}
  \memoize(\mktab, ((\mexpr,\menv), \mstore'), \mathit{v\sigma}) &=
  [((\mexpr,\menv), \mstore') \mapsto \set{{\mathit{v\sigma}}}] \\
  \memoize(\mktab, ((\mexpr,\menv), \maddr), \mathit{v\sigma}) &=
  \bigsqcup\limits_{\mstore' \in \mktab(\maddr)}\snglm{((\mexpr,\menv), \mstore')}{\mathit{v\sigma}}
\end{align*}
At capture time, we strip the store in the $\rt$ continuation if there is one, and replace it with an address:

@@

The immediately related work is that of PDCFA \citep{ianjohnson:earl2010pdcfa}, CFA2~\citep{ianjohnson:vardoulakis-lmcs11, ianjohnson:Vardoulakis2011Pushdown}, and AAM ~\citep{dvanhorn:VanHorn2010Abstracting}, the first two of which we recreated in full detail.
%
The version of CFA2 that handles \rackett{call/cc} does not handle composable control, and is dependent on a restricted CPS representation.
%
They also had no way to tune the precision of their first class continuation approximation.
%
Since \rackett{call/cc} can be simulated with shift/reset, this work supercedes theirs.
%
The extended version of PDCFA that does garbage collection~\citep{dvanhorn:Earl2012Introspective} also fits into our model;
the addresses that the stack keeps alive can be accumulated by ``reading through'' the continuation table, building up the set of addresses in each portion of the stack that we come across.
%%

>>>>

The $\Addr$ space is what controls the precision of the model.
%
For a concrete semantics, we require the allocation meta-function $\alloc :
\State \to \Addr$ to return fresh addresses ($\alloc(\mstate) \notin \dom(\mstate.\mstore)$), but any choice of address is sound~\citep{dvanhorn:Might2009Posteriori}.
%
If $\alloc$ only uses addresses from a finite subset of $\Addr$, then the state space without continuations is finite, and the space of contination frames is finite (thus the pushdown system interpretation is apt).
%
This meta-function along with the commitment to having no recursive data-structures is the key to the AAM technique.
%
All recursion can be expressed by indirecting through addresses into a store --- tying Landin's knot.
%
We borrow this technique on top of our own.

@@

To turn this semantics into PDCFA's algorithm for constructing a Dyck state graph, given the appropriate $\alloc$, we apply a widening operator to make $\mstore$, $\mkont$, and $\mmemo$ shared amongst all states (meaning states then become represented without these components), in what is then called a $\System$.
%
Since these components are shared, they are always the least upper bound of any respective component that the semantics produces (the intermediate set $I$) in order to stay sound.
%
We use a meta-function $\wn$, ``wide to narrow'', to shift between the two representations of states.

%% FIXME: remove stores from memo-table since they will always be unchanged
%% TODO: Show that store widening and store/memo/KTab widening have a bisimulation.
\begin{align*}
  \mastate \in \sa{State} &= (\Expr \times \Env) \times \Kont \\
  \System &= \wp(\sa{State} \times \Store) \times \wp(\sa{State}) \times \Store \times \KTab \times \Memo \\
  {\mathcal F} &: \sa{\State} \to \System \to \System \\
  {\mathcal F}(\mstate_0)(S, F, \mstore, \mktab, \mmemo) &= (\set{(\mstate_0,\bot)} \cup S \cup S', F', \mstore', \mktab', \mmemo') \\
  \text{where } I &= \setbuild{\mstate'}{\mastate \in F, \wn(\mastate,\mstore,\mktab,\mmemo) \stepto \mstate'} \\
                \mstore' &= \bigsqcup\setbuild{\mstore'}{\wn(\_, \mstore',\_,\_) \in I} \\
                \mktab' &=  \bigsqcup\setbuild{\mktab'}{\wn(\_, \_, \mktab', \_) \in I} \\
                \mmemo' &=  \bigsqcup\setbuild{\mmemo'}{\wn(\_, \_, \_, \mmemo') \in I} \\
                S' &= \setbuild{(\mastate',\mstore')}{\wn(\mastate', \_, \_, \_) \in I} \\
                F' &= \setbuild{\mastate'}{(\mastate',\_) \in S' \setminus S} \\
                \wn(\tpl{(\mexpr,\menv),\mkont},\mstore,\mktab,\mmemo)
                   &= \tpl{(\mexpr,\menv),\mstore,\mkont,\mktab,\mmemo}
\end{align*}

@@

The different tables we have encode information about execution history that we prove is invariant.
\begin{align*}
  \inv(\tpl{\mpoint, \mstore, \mkont, \mktab, \mmemo}) &=
          \dom(\mmemo) \subseteq \dom(\mktab) \\
  &\wedge \forall (\mpoint',\mstore') \in \dom(\mktab),
                  \mkont^\circ \in \mktab(\mpoint',\mstore'),
                  \mkont' \in \unroll{\mktab}{\mkont^\circ}. \\   
  &\qquad \phi_{\reachable} \wedge \phi_{\memo} \\
  \text{where } \phi_{\reachable} &=
  \exists \mkont^* \in \unroll{\mktab}{\mkont}.
    \tpl{\mpoint', \mstore', \mkont'} \stepto^* \tpl{\mpoint, \mstore, \mkont^*} \\
  \phi_{\memo} &=
 \forall (\mval, \mstore'') \in \mmemo(\mpoint',\mstore'). \\
  &\qquad \exists \mtrace\equiv\tpl{\mpoint', \mstore', \mkont'} \stepto^* \tpl{\mval,\mstore'', \mkont'}.
     \hastail(\mtrace,\mkont') \\
\end{align*}

The first is that the memo table only contains information about previously seen contexts; we need this to infer that there was at least one call leading to the memoized context such that we can use stack irrelevance to justify skipping to the memoized result.
%
The second, $\phi_{\reachable}$, states that all the calling contexts in the continuation table reach some unrolling of the current state.
%
The final invariant, $\phi_{\memo}$, states that the memo table is correct: all calling contexts reach the memoized result, with the calling context's continuation in the tail (to justify using stack-irrelevance).

@@

This semantics of course enjoys the soundness properties that GC does not introduce dangling pointers (\autoref{thm:gc-pointers}), equal length traces are equal up to GC (in the concrete) (\autoref{thm:gc-concrete}), and abstractions are sound with respect to collected states (\autoref{thm:gc-sound}).
%
\begin{theorem}\label{thm:gc-pointers}
  All addresses in $\Gamma(\mstate)$ are in the domain of $\Gamma(\mstate).\mstore$.
\end{theorem}

\begin{theorem}\label{thm:gc-concrete}
  If $\mstate \stepto^n \mstate'$ and $\mstate \stepto^n \mstate''$ then $\Gamma(\mstate') = \Gamma(\mstate'')$.
\end{theorem}

\begin{theorem}\label{thm:gc-sound}
  If $\mstate \stepto \mstate'$ and $\alpha(\Gamma(\mstate)) \sqsubseteq \mastate$ then there is a $\mastate'$ such that $\mastate \astepto \mastate'$ and $\alpha(\Gamma(\mstate')) \sqsubseteq \mastate'$.
\end{theorem}
%
Unconstrained allocation functions

@@

\begin{align*}
  \msctx \in \SContext &::= ((\mexpr,\menv),\maddr) \\
  \mctx \in \Context &::= \msctx \mid \ldots \\
  \mskont \in \SKont &= \krt{\msctx} \mid \ldots \\
  \mval \in \Value &= \vclo{\slam{\mvar}{\mexpr}}{\menv} \mid \vcomp{\mskont} \\
  \mktab \in \KTab &= (\Context \to \mathit{KTabRange}) \\
                   &\cup (\Addr \to \wp(\Store))
\end{align*}

@@

At return boundaries, the memo table must add the result to all the represented contexts:
\begin{align*}
  \memoize(\mktab, ((\mexpr,\menv), \mstore'), \mathit{v\sigma}) &=
  [((\mexpr,\menv), \mstore') \mapsto \set{{\mathit{v\sigma}}}] \\
  \memoize(\mktab, ((\mexpr,\menv), \maddr), \mathit{v\sigma}) &=
  \bigsqcup\limits_{\mstore' \in \mktab(\maddr)}\snglm{((\mexpr,\menv), \mstore')}{\mathit{v\sigma}}
\end{align*}
At capture time, we strip the store in the $\rt$ continuation if there is one, and replace it with an address:

@@

The immediately related work is that of PDCFA \citep{ianjohnson:earl2010pdcfa}, CFA2~\citep{ianjohnson:vardoulakis-lmcs11, ianjohnson:Vardoulakis2011Pushdown}, and AAM ~\citep{dvanhorn:VanHorn2010Abstracting}, the first two of which we recreated in full detail.
%
The version of CFA2 that handles \rackett{call/cc} does not handle composable control, and is dependent on a restricted CPS representation.
%
They also had no way to tune the precision of their first class continuation approximation.
%
Since \rackett{call/cc} can be simulated with shift/reset\footnote{\RktPn{(}\RktSym{define}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{call/cc}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{f}\RktPn{)}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{shift}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{k}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{k}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{f}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{$\lambda$}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{x}\RktPn{)}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{shift}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{c}\mbox{\hphantom{\Scribtexttt{x}}}\RktPn{(}\RktSym{k}\mbox{\hphantom{\Scribtexttt{x}}}\RktSym{x}\RktPn{)}\RktPn{)}\RktPn{)}\RktPn{)}\RktPn{)}\RktPn{)}\RktPn{)}}, this work supercedes theirs (the space complexity difference imposed by the simulation is non-issue in the abstract).
%
The extended version of PDCFA that does garbage collection~\citep{dvanhorn:Earl2012Introspective} also fits into our model;
the addresses that the stack keeps alive can be accumulated by ``reading through'' the continuation table, building up the set of addresses in each portion of the stack that we come across.
%%
